Diversity dynamics in Nymphalidae butterflies: Effect of phylogenetic uncertainty on diversification rate shift estimates

### 2014-11-11 MS_revised.tex
adding BAMM results as fig S01

We contrasted these results with those obtained from analyzing the data in
the software BAMM \cite{rabosky2014}. We found 
that the best model (best configuration) estimated 2 diversification shifts in
our MCC tree (posterior probability = 0.79) for the Ithomiini
 and Satyrini. Bayes factor (BF) values strongly 
favored this model in comparison with a constant rate model (no diversification
shifts, BF $>$ 1,884).

@@

The MEDUSA analyses, taking into account phylogenetic uncertainty,
estimated a net diversification rate increase in part of the clade
Ithomiini across more than 95\% of the trees. Our BiSSE analysis found a
positive effect of the character state ``feeding on Solanaceae'' on the
net net diversification rate on part of Ithomiini (Oleriina + Ithomiina
+ Napeogenina + Dircennina + Godyridina) (Fig. 4). The estimated mean
net diversification rate for taxa that do not feed on Solanaceae was
\texttt{r = 0.11} while the net diversification rate for the Solanaceae
feeders was \texttt{r = 0.16} (see Fig. S1 for a boxplot of speciation
and extinction values for the 95\% credibility intervals). The same
analysis considering \emph{Vanessa} and \emph{Hypanartia} as
non-Solanaceae feeders due to dubious records produced the same pattern
and net diversification rates (Fig. S2). Therefore, the rest of the
analyses were performed assuming these two genera as Solanaceae feeders.
We constrained the BiSSE likelihood model to force equal rates of
speciation for both character states in order to test whether the model
of different speciation rates is a significantly better explanation for
the data. A likelihood ratio test found that the model for increased net
diversification rate for nymphalids feeding on Solanaceae is a
significantly better explanation than this character state having no
effect on diversification (\(\chi^2 = 12.3; 1 df; p < 0.001\)) (Table 3;
character states available in supp. mat. 17, code in supp. mat. 18, and
mcmc run in supp. mat. 19). We combined the post-burnin mcmc generations
from running BiSSE on 250 trees from the random sample of 1000 trees and
found the same pattern as the BiSSE analysis on the maximum clade
credibility tree (combined mcmc run in supp. mat. 20; profiles plot of
speciation rates in Fig. S3; boxplot of 95\% credibility intervals in
Fig. S4). A BiSSE analysis to test whether the trait ``feeding on
Apocynaceae'' had any effect on increased net diversification rates
found similar speciation rates for lineages feeding on Apocynaceae and
other plants (Fig. S5).

@@

 \item {\bf Figure S1.} Boxplot of speciation ($\lambda$) and extinction
($\mu$) values for the 95\% credibility intervals of values estimated
by BiSSE analysis of diversification due to feeding on Solanaceae
plants.

 \item {\bf Figure S2.} BiSSE analysis of diversification of nymphalids due
to feeding on Solanaceae hostplants assuming \emph{Vanessa} and
\emph{Hypanartia} as non-Solanaceae feeders. The same pattern is
recovered, speciation and net diversification rates are significantly
higher for Solanaceae feeders $\lambda1$, r1).

 \item {\bf Figure S3.} Net diversification rates of nymphalids feeding on
Solanaceae plants as estimated by combining post-burnin runs of BiSSE on
the 1000 trees from the posterior distribution.

 \item {\bf Figure S4.} Boxplot of speciation and extinction values for the
95\% credibility intervals of values estimated by BiSSE analysis of
diversification due to feeding on Solanaceae plants on the combined
post-burnin runs on 1000 trees from the posterior distribution.

 \item {\bf Figure S5.} BiSSE analysis of diversification of nymphalids due
to feeding on Apocynaceae hostplants. Speciation and net diversification
rates are similar.
\end{description}

>>>>

We contrasted these results with those obtained from analyzing the data in
the software BAMM \cite{rabosky2014}. We found 
that the best model (best configuration) estimated 2 diversification shifts in
our MCC tree (posterior probability = 0.79) for the Ithomiini
and Satyrini (Fig. S1). Bayes factor (BF) values strongly 
favored this model in comparison with a constant rate model (no diversification
shifts, BF $>$ 1,884).

@@

The MEDUSA analyses, taking into account phylogenetic uncertainty,
estimated a net diversification rate increase in part of the clade
Ithomiini across more than 95\% of the trees. Our BiSSE analysis found a
positive effect of the character state ``feeding on Solanaceae'' on the
net net diversification rate on part of Ithomiini (Oleriina + Ithomiina
+ Napeogenina + Dircennina + Godyridina) (Fig. 4). The estimated mean
net diversification rate for taxa that do not feed on Solanaceae was
\texttt{r = 0.11} while the net diversification rate for the Solanaceae
feeders was \texttt{r = 0.16} (see Fig. S2 for a boxplot of speciation
and extinction values for the 95\% credibility intervals). The same
analysis considering \emph{Vanessa} and \emph{Hypanartia} as
non-Solanaceae feeders due to dubious records produced the same pattern
and net diversification rates (Fig. S3). Therefore, the rest of the
analyses were performed assuming these two genera as Solanaceae feeders.
We constrained the BiSSE likelihood model to force equal rates of
speciation for both character states in order to test whether the model
of different speciation rates is a significantly better explanation for
the data. A likelihood ratio test found that the model for increased net
diversification rate for nymphalids feeding on Solanaceae is a
significantly better explanation than this character state having no
effect on diversification (\(\chi^2 = 12.3; 1 df; p < 0.001\)) (Table 3;
character states available in supp. mat. 17, code in supp. mat. 18, and
mcmc run in supp. mat. 19). We combined the post-burnin mcmc generations
from running BiSSE on 250 trees from the random sample of 1000 trees and
found the same pattern as the BiSSE analysis on the maximum clade
credibility tree (combined mcmc run in supp. mat. 20; profiles plot of
speciation rates in Fig. S4; boxplot of 95\% credibility intervals in
Fig. S5). A BiSSE analysis to test whether the trait ``feeding on
Apocynaceae'' had any effect on increased net diversification rates
found similar speciation rates for lineages feeding on Apocynaceae and
other plants (Fig. S6).

@@

 \item {\bf Figure S1.} Best diversification model as estimated by BAMM. Two
diversification shifts were recovered (Ithomiini and Satyrini). The MEDUSA 
analysis found a diversification shifts in Satyrinae (the parent node of 
Satyini).

 \item {\bf Figure S2.} Boxplot of speciation ($\lambda$) and extinction
($\mu$) values for the 95\% credibility intervals of values estimated
by BiSSE analysis of diversification due to feeding on Solanaceae
plants.

 \item {\bf Figure S3.} BiSSE analysis of diversification of nymphalids due
to feeding on Solanaceae hostplants assuming \emph{Vanessa} and
\emph{Hypanartia} as non-Solanaceae feeders. The same pattern is
recovered, speciation and net diversification rates are significantly
higher for Solanaceae feeders $\lambda1$, r1).

 \item {\bf Figure S4.} Net diversification rates of nymphalids feeding on
Solanaceae plants as estimated by combining post-burnin runs of BiSSE on
the 1000 trees from the posterior distribution.

 \item {\bf Figure S5.} Boxplot of speciation and extinction values for the
95\% credibility intervals of values estimated by BiSSE analysis of
diversification due to feeding on Solanaceae plants on the combined
post-burnin runs on 1000 trees from the posterior distribution.

 \item {\bf Figure S6.} BiSSE analysis of diversification of nymphalids due
to feeding on Apocynaceae hostplants. Speciation and net diversification
rates are similar.
\end{description}
